---
title: "Poisson Regression"
author: "Anubhav Dhiman"
date: "3/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What Are Poisson Regression Models?
Poisson Regression models are best used for modeling events where the outcomes are counts

## What is Poisson Distribution?
It models the probability of event or events $y$ ocurring within a specific timeframe, assuming that $y$ occurences are not affected by the timing of previous ocurrences of $y$.

```{r}
colors<-c("Red", "Blue", "Gold", "Black", "Pink", "Green") #vector of colors
poisson.dist<-list() #declare a list to hold distribution values
a<-c(1, 2, 3, 4, 5, 6) # A vector for values of u

for (i in 1:6) {
  poisson.dist[[i]]<-c(dpois(0:20, i)) 
  # Store distribution vector for each corresonding value of u
}

#plot each vector in the list using the colors vectors to represent each value for u
 plot(unlist(poisson.dist[1]), type = "o", xlab="y", ylab = "P(y)", col=colors[i])

for (i in 1:6) {
 lines(unlist(poisson.dist[i]), type = "o", col=colors[i])
  
}
 # Adds legend to the graph plotted
legend("topright", legend=a, inset= 0.08,  cex = 1.0, fill = colors, title = "Values of u")
```

## How Does Poisson Distribution Differ From Normal Distribution?
Poisson Distribution is most commonly used to find the probability of events occurring within a given time interval. Since we're talking about a count, with Poisson distribution, the result must be 0 or higher - it's not possible for an event to happen a negative number of times. On the other hand, Normal distribution is a continuous distribution for a continuous variable and it could result in a positive or negative value

```{r}
#create a sequence -3 to +3 with .05 increments 
xseq<-seq(-3, 3, .05) 

#generate a Probability Density Function
densities<-dnorm(xseq, 0, 1) 

#plot the graph
plot(xseq, densities, col="blue", xlab="", ylab="Density", type="l", lwd=2) 
# col: changes the color of line
# xlab: and 'ylab' are labels for x and y axis respectively
# type: defines the type of plot. 'l' gives a line graph
# lwd: defines line width
```

### If there are 12 cars crossing a bridge per minute on average, what is the probability of having seventeen or more cars crossing the bridge in any given minute?

```{r}
# u = 12
ppois(16, 12, lower.tail = FALSE)
#lower.tail=logical; if TRUE (default) then probabilities are P[X<=x], otherwise, P[X > x].
```

## Poisson Regression Models and GLMs
Generalized Linear Models are models in which response variables follow a distribution other than the normal distribution. That's in contrast to Linear regression models, in which response variables follow normal distribution.

A Poisson Regression model is a Generalized Linear Model (GLM) that is used to model count data and contingency tables. The output $Y$ (count) is a value that follows the Poisson distribution. It assumes the logarithm of expected values (mean) that can be modeled into a linear form by some unknown parameters.

To transform the non-linear relationship to linear form, a link function is used which is the log for Poisson Regression. For that reason, a Poisson Regression model is also called **log-linear model**.

The coefficients are calculated using methods such as Maximum Likelihood Estimation(MLE) or maximum quasi-likelihood.

One of the most important characteristics for Poisson distribution and Poisson Regression is *equidispersion*, which means that the mean and variance of the distribution are equal.

Let us say that the mean ($\mu$) is denoted by $E(X)$. 
$E(X)=\mu$

For Poisson Regression, mean and variance are related as:
$var(X)=\sigma^2E(X)$

Where $\sigma^2$ is the dispersion parameter. Since $var(X)=E(X)$ (variance=mean) must hold for the Poisson model to be completely fit, $\sigma^2$ must be equal to $1$.

When variance is greater than mean, that is called **over-dispersion** and it is greater than 1. If it is less than 1 than it is known as **under-dispersion**.

## Poisson Regression Modeling Using Count Data

We're going to model Poisson Regression related to how frequently yarn breaks during weaving.

```{r}
#install.packages("datasets")
library(datasets) #include library datasets after installation

data<-warpbreaks
columns<-names(data) #Extract column names from dataframe
columns #show columns

ls.str(warpbreaks)

hist(data$breaks) 
#generate histogram of new.data$whrswk

mean(data$breaks) #calculate mean

var(data$breaks) #calculate variance
```

The variance is much greater than the mean, which suggests that we will have over-dispersion in the model.

```{r}
#model poisson regression usin glm()
poisson.model<-glm(breaks~wool+tension, data, family = poisson(link = "log"))

summary(poisson.model)
```
If the p is less than 0.05 then, the variable has an effect on the response variable. In the summary above, we can see that all p values are less than 0.05, hence, both explanatory variables (wool and tension) have significant effect on breaks.

Before starting to interpret results, let's check whether the model has over-dispersion or under-dispersion. If the Residual Deviance is greater than the degrees of freedom, then over-dispersion exists. This means that the estimates are correct, but the standard errors (standard deviation) are wrong and unaccounted for by the model.

The Null deviance shows how well the response variable is predicted by a model that includes only the intercept (grand mean) whereas residual with the inclusion of independent variables. Above, we can see that the addition of 3 (53-50 =3) independent variables decreased the deviance to 210.39 from 297.37. Greater difference in values means a bad fit.

So, to have a more correct standard error we can use a quasi-poisson model:

```{r}
poisson.model2<-glm(breaks~wool+tension, data = data, family = quasipoisson(link = "log"))

summary(poisson.model)
```

### Comparing The Models

```{r}
#install.packages("arm")
# load library arm that contains the function se.coef()
library(arm) 

#extract co-effcients from first model using 'coef()'
coef1=coef(poisson.model) 

#extract co-effcients from second model
coef2=coef(poisson.model2) 

#extract standard errors from first model using 'se.coef()'
se.coef1=se.coef(poisson.model) 

#extract standard errors from second model
se.coef2=se.coef(poisson.model2)

#use 'cbind()' to combine values one dataframe
models.both<-cbind(coef1, se.coef1, coef2, se.coef2, exponent=exp(coef1)) 

# show dataframe
models.both 
```

In above output, we can see the coefficients are the same, but the standard errors are different.

Keeping these points in mind, let's see estimate for *wool*.
```{r}
coeff_wool = exp(-0.2059884)
1-coeff_wool
```
If we change wool type from A to B, the number of breaks will fall by 18.6% assuming all other variables are the same.

### Predicting From The Model
```{r}
#make a datframe with new data
newdata = data.frame(wool="B", tension="M")

#use 'predict() to run model on new data
predict(poisson.model2, newdata = newdata, type = "response")
```

Our model is predicting there will be roughly 24 breaks with wool type B and tension level M.

## Visualizing Findings

```{r}
#Install the package jtools if not already installed
#install.packages("jtools")

#you may be asked to install 'broom' and 'ggstance' packages as well
#install.packages("broom")
#install.packages("ggstance")

#Include jtools library
library(jtools)

#plot regression coefficients for poisson.model2
plot_summs(poisson.model2, scale = TRUE, exp = TRUE)
```

```{r}
#plot regression coefficients for poisson.model2 and poisson.model
plot_summs(poisson.model, poisson.model2, scale = TRUE, exp = TRUE)
```

```{r}
# install.packages("interactions")
cat_plot(poisson.model2, pred = wool, modx = tension)
#argument 1=regression model
#pred=The categorical variable that will appear on x-axis
#modx=A categorical moderator variable. Moderator variable that has an effect in combination to pred on outcome
```

```{r}
# using cat_plot. Pass poisson.model2 and we want to see effect of tension type so we set pred=tension
cat_plot(poisson.model2, pred = tension, modx=wool)
```

```{r}
cat_plot(poisson.model2, pred = tension, modx = wool, geom = "line")
```

```{r}
cat_plot(poisson.model2, pred = tension, modx = wool, geom = "line", plot.points = TRUE)
```

## Poisson Regression Modeling Using Rate Data
So far we have modeled count data, but we can also model rate data that is predicting the number of counts over a period of time or grouping. 

```{r}
#install.packages("ISwR")
library(ISwR)

data(eba1977)
cancer.data = eba1977
cancer.data[1:10,]
#Description
#Lung cancer incidence in four Danish cities 1968-1971

#Description:
#     This data set contains counts of incident lung cancer cases and
#     population size in four neighbouring Danish cities by age group.

#Format:
#     A data frame with 24 observations on the following 4 variables:
#     city a factor with levels Fredericia, Horsens, Kolding, and Vejle.
#     age a factor with levels 40-54, 55-59, 60-64, 65-69,70-74, and 75+.
#     pop a numeric vector, number of inhabitants.
#    cases a numeric vector, number of lung cancer cases.
```

```{r}
#find the log(n) of each value in 'pop' column. It is the third column
logpop = log(cancer.data[,3])

#add the log values to the dataframe using 'cbind()'
new.cancer.data = cbind(cancer.data, logpop)

#display new dataframe
new.cancer.data
```
```{r}
poisson.model.rate <- glm(cases ~ city + age+ offset(logpop), family = poisson(link = "log"), data = cancer.data)

#display summary
summary(poisson.model.rate)
```

In this dataset, we can see that the residual deviance is near to degrees of freedom, and the dispersion parameter is 1.5 (23.447/15) which is small, so the model is a good fit.

```{r}
fitted(poisson.model.rate)
```

```{r}
#create a test dataframe containing new values of variables
test.data=data.frame(city = "Kolding", age = "40-54", pop = 1000, logpop = log(1000)) 

#predict outcomes (responses) using 'predict()' 
predicted.value<-predict(poisson.model.rate, test.data, type = "response") 

#show predicted value
predicted.value
```

So, for the city of Kolding among people in the age group 40-54, we could expect roughly 2 or 3 cases of lung cancer per 1000 people.

As with the count data, we could also use quasi-poisson to get more correct standard errors with rate data.



